---
title: "Assignment1-AdvRprogramming"
author: "Vishweshwar Chowdhury; 24205804"
format:
  revealjs:
    slide-number: true
    html-math-method: mathjax
    theme: serif
    footer: ""
    incremental: false
    chalkboard: false
    menu: true
    scrollable: true
    preview-links: false
    embed-resources: true
    standalone: true
    transition: convex
execute:
  echo: true
editor: visual
---

## Introduction

This project analyzes key development indicators for **France, Kazakhstan, and Ukraine** using World Bank data. We explore trends across areas like health, education, economy, and environment using the data.table package in R.

```{r, echo=FALSE}
#Loading required Libraries
library(data.table)
library(ggplot2)
library(scales)
```

## 1) Read Data using data.table and assign correct class to variables {.smaller}

```{r}
# Reading each country's data
fra <- fread("indicators_fra.csv")
kaz <- fread("indicators_kaz.csv")
ukr <- fread("indicators_ukr.csv")
options(datatable.print.topn = 3,datatable.print.nrows = 5,width = 80)       
str(fra)
head(fra)
options(datatable.print.topn = 3,datatable.print.nrows = 5,width = 80)       
str(kaz)
head(kaz)
options(datatable.print.topn = 3,datatable.print.nrows = 5,width = 80)       
str(ukr)
head(ukr)
```

## In the above output we can observe the following things:-

-   Each dataset had 6 columns: Country Name, Country ISO3, Year, Indicator Name, Indicator Code, Value.

-   The first row was incorrectly read as data (e.g., #country+name), so it was removed during cleaning. All columns were initially read as character; Year and Value were later converted to integer and numeric respectively.

## 2) Merging Datasets using data.table

```{r}
# Add a country label to each dataset
fra[, Country := "France"]
kaz[, Country := "Kazakhstan"]
ukr[, Country := "Ukraine"]

# Combine the three datasets into one
all_data <- rbindlist(list(fra, kaz, ukr),
use.names = TRUE, fill = TRUE)

# Check structure and preview
str(all_data)
head(all_data)
```

## The merged dataset contains the following things:-

-   The merged dataset contains **222,244 rows** and 7 variables.

-   Column names and structure are consistent across all countries.

-   The resulting dataset allows for cross-country comparisons across years and indicators.

## 3) Data Exploration

```{r}
# Basic summary
summary(all_data)
# Number of unique indicators
length(unique(all_data$`Indicator Name`))
# Time coverage
range(all_data$Year, na.rm = TRUE)
```

## 

```{r}
# Count of rows per country
all_data[, .N, by = Country]
# Top 5 most common indicators
all_data[, .N, by = `Indicator Name`][order(-N)][1:5]
# Missing values check
colSums(is.na(all_data))
# Number of observations per year per country
all_data[, .N, by = .(Country, Year)][order(Country, Year)]
```

## The findings of the data analysis task are as follows:-

-   The merged dataset has **222,244 rows with 7 character columns**, indicating raw data still needed cleaning.

-   Indicators are diverse, with the most common including:

    -   **Net migration**

    -   **Life expectancy**

    -   **Adolescent fertility rate**

    -   **Under 5-mortality rate**

-   The dataset spans a wide time range — **from 1960 to 2023** - with varying data density per year.

-   France had data available for nearly all years between **1960 and 2023**, with some variation in the number of observations per year.

## 4) Data Analysis task using keyby argument

```{r}
# Viewing the most common indicators
all_data[, .N, by = `Indicator Name`][order(-N)][1:20]
# Filter for selected indicators
focus_indicators <- c("Mortality rate, under-5 (per 1,000 live births)",
                      "Net migration",
                      "Population ages 15-64 (% of total population)")

selected_data <- all_data[`Indicator Name` %in% focus_indicators]
selected_data[, Value := as.numeric(Value)]
# Mean values per indicator per country over time
summary_data <- selected_data[, .(Average = mean(Value, na.rm = TRUE)),
                              keyby = .(Country, `Indicator Name`, Year)]

head(summary_data)
```

## The following observations were made from above output:-

1.  Using data.table 5 relevant development indicators were found which are net migration,Adolescent fertility rate , Life expectancy at birth and mortality rate under 5.

2.  The data was grouped and summarized by Country, Indicator Name, and Year using **keyby** to calculate yearly averages.

## 5) Creation of Development Indicators over time plot and Under 5 Mortality Rate over time plot

## Code

```{r}
# Clean Year column
all_data <- all_data[!is.na(Year)]             
all_data <- all_data[Year != "#date+year"]

# Convert Year to numeric
all_data[, Year := as.integer(Year)]

# Filter indicators
indicators_to_plot <- c(
  "Net migration",
  "Adolescent fertility rate (births per 1,000 women ages 15-19)",
  "Life expectancy at birth, female (years)",
  "Life expectancy at birth, male (years)",
  "Mortality rate, under-5 (per 1,000 live births)"
)

plot_data <- all_data[`Indicator Name` %in% indicators_to_plot]
plot_data[, Value := as.numeric(Value)]

# Recalculate summaries
summary_data <- plot_data[
  , .(Average = mean(Value, na.rm = TRUE)),
  keyby = .(Country, `Indicator Name`, Year)
]
```

## Creation of Development Indicators over time- Plot

```{r, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE, message=FALSE}
ggplot(summary_data, aes(x = Year, y = Average, color = Country)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~ `Indicator Name`, 
  scales = "free", ncol = 2, 
  labeller = label_wrap_gen(width = 30)) +
  labs(title = "Development Indicators Over Time",
       y = "Average Value", x = "Year") +
  scale_y_continuous(labels = label_comma()) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(size = 13, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

## Under 5 mortality rate over time- Plot

```{r, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE, message=FALSE}
mortality_data <- summary_data[
  `Indicator Name` == "Mortality rate, under-5 (per 1,000 live births)"
]
ggplot(mortality_data, aes(x = Year, y = Average, color = Country)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Under-5 Mortality Rate Over Time",
       y = "Deaths per 1,000 Live Births", x = "Year") +
  theme_minimal()
```

## [Plot interpretations]{.underline} {.center}

## 1) Development Indicators over time

-   **Life expectancy (male and female)** shows consistent growth across all countries, with **France** leading, followed by Ukraine and Kazakhstan.

-   **Under-5 mortality rates** have significantly declined in all three countries — reflecting healthcare improvements — **though Kazakhstan started from a much higher rate**.

-   **Adolescent fertility rates** have steadily **declined**, **especially in France**, indicating improved reproductive health awareness.

-   **Net migration** shows **extreme reduction in Ukraine,** likely reflecting the recent **war** between Ukraine and palestine.

## 2) Under-5 Mortality Rate Over Time

-   **Kazakhstan** had the **highest child mortality in the 1960s**, but saw major declines over time.

-   **France** maintained the **lowest under-5 mortality rates throughout the timeline,** reflecting a consistently **strong public health system**.

-   **Ukraine** showed moderate improvement, with a **steady decline from the 1970s to present.**

-   Overall, the plot demonstrates strong downward trends for all three countries, **emphasizing global** **progress in reducing child mortality.**
