---
title: "Adv R Program HW 1"
author: "Aryan"
format: pdf
editor: visual
execute:
  warning: false  
  error: false    
  message: false
  echo: false
---

# Homework 1

### [*Introduction*]{.underline}

***Volcano*** is a `numeric` matrix, inbuilt R dataset. The dataset represents *`Topographic Information on Auckland's Maunga Whau Volcano`*

```{r}
data("volcano")
```

```{r}
str(volcano)
```

---

### [*About Data*]{.underline}

-   The data contains as a *matrix* with:

    -   **87 rows** (representing Y-coordinates)

    -   **61 columns** (representing X-coordinates)

-   Each cell contains an **elevation value** (in *`meters`*).

#### [Interpretation of the findings]{.underline}

-   This is the **lowest elevation** in the dataset is `94m`.

-   `25%` of the elevation values are **less than or equal to `108` meters** .

-   Half of the elevation values are **below `124`** , and half are **above `124`** .

-   the **average elevation** across all points in the data is `130`.

-   `75%` of the elevation values are **less than or equal to `150` meters**.

-   A standard deviation of **`25.83`** indicating a **moderate variability** in elevation.

```{r}
vec = as.vector(volcano)

stats = c(
  Mini = min(vec),
  "1st Qt" = quantile(vec, 0.25),
  Median = median(vec),
  Mean = mean(vec),
  "3rd Qt" = quantile(vec, 0.75),
  "Range" = max(vec) - min(vec),
  "Stand Dev" = sd(vec)
)

round(stats, 2)
```

---

### [*Image Plot of the topological information*]{.underline}

```{r}
#Image plot
image(volcano, 
      col = terrain.colors(255), #for gradient of colors- green(low) and orange(high)
      main= "Topographic Information on Auckland's Maunga Whau Volcano")
box()
contour(volcano, add = TRUE, col = "white")
```

#### [Interpretation of the graph]{.underline}

-   The color gradient ranges from ***green to yellow to orange***, representing different elevation levels.

    -   **Green areas** indicate *`lower elevations`*.

    -   **Yellow** indicates *`medium elevations`*.

    -   **orange areas** indicate *`higher elevations`*.

-   The `gradient` helps quickly identify regions of `varying heights` across the volcano.

---

### Reference

\[1\] Image() function in [Rdocumentation](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/image).

\[2\] Contour function in [Rdocumentation](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/contour).
